---
title: 'Intro Session 1: Motivating reproducibility'
author: ''
date: ''
output:
  html_document:
    fig_height: 3
    fig_width: 6
---

This RMarkdown file contains R code you can use to complete the exercises from Intro Session 1. To see the output simply click on Knit HTML above. This might prompt you to install and load some required packages, specifically `knitr`. Just click yes, and the a document including this narrative, the R code, and figures should pop up.

## Data: 

Load data using the `read.csv` function:

```{r}
gapminder_5060 = read.csv("data/gapminder-5060.csv")
```

## Task 1:

Visualize the relationship between life expectancy over time for Canada in the 1950s and 1960s using a line plot.

* Create a subset of `gapminder_5060` for Canada:
```{r}
ca_5060 = gapminder_5060[gapminder_5060$country == "Canada", ]
```

* Next we will create the plot, but in order to do so we will use the `ggplot2` package. This is not the only way to make this plot, but the `ggplot2` package has some aesthetic defaults that makes it attractive:

```{r}
# check to see if the ggplot2 package is already installed, and if not, install it
if (!("ggplot2" %in% names(installed.packages()[,"Package"]))) {install.packages("ggplot2")}

# load the ggplot2 package
library(ggplot2)
```

Now we make the plot:
```{r}
ggplot(data = ca_5060, aes(x = year, y = lifeExp)) +
  geom_point() +
  geom_line() 
```
We will talk 

### Task 1 - Stretch goal:

Add lines for Mexico and US in the 1950s and 1960s.

* Create a subset of `gapminder_5060` for Canada, Mexico, and US:
```{r}
countries = c("Canada", "Mexico", "United States")
northamer_5060 = gapminder_5060[gapminder_5060$country %in% countries, ]
```

* Create the scatterplot using `ggplot2`:
```{r}
library(ggplot2)
ggplot(data = northamer_5060, aes(x = year, y = lifeExp, colour = country)) +
  geom_line() +
  geom_point()
```

## Task 2:

Visualize the relationship between GDP and life expectancy for countries in Europe in 1952.

* Create a subset of `gapminder_5060` for Europe in 1952:
```{r}
eu_52 = gapminder_5060[gapminder_5060$continent == "Europe" & gapminder_5060$year == 1952, ]
```

* Create the plot:
```{r}
ggplot(data = eu_52, aes(x = lifeExp, y = gdpPercap)) +
  geom_line() +
  geom_point()
```

### Task 2 - Stretch goal:

Add year 1967 in another color.

* Create a subset of `gapminder_5060` for Europe in 1952 and 1967:
```{r}
eu_5267 = gapminder_5060[gapminder_5060$continent == "Europe" & (gapminder_5060$year == 1952 | gapminder_5060$year == 1967), ]
```

* Create the plot:
```{r}
ggplot(data = eu_5267, aes(x = lifeExp, y = gdpPercap, colour = as.factor(year))) +
  geom_line() +
  geom_point()
```