---
title: 'Intro Session 1: Motivating reproducibility'
author: ''
date: ''
output:
  html_document:
    fig_height: 3
    fig_width: 6
---

This RMarkdown file contains R code you can use to complete the exercises from Intro Session 1. To see the output simply click on Knit HTML above. This might prompt you to install and load some required packages, specifically `knitr`. Just click yes, and the a document including this narrative, the R code, and figures should pop up.

## Data: 

The `read.csv` function is used to read the data into R. Note that the argument provided for this function is the complete path that leads to the dataset from your current working directory (where this RMarkdown file is located). Also note that this is provided as a character string, hence in quotation marks.

```{r}
gapminder_5060 = read.csv("data/gapminder-5060.csv")
```

## Packages:

We will use the `ggplot2` package to make the plots for these exercises. This is not the only way to make this plot, but the `ggplot2` package has some aesthetic defaults that makes it attractive. 

The first time you need a package you need to first download and install it. The function below checks to see if the package is already installed, and if not it installs it.

```{r}
if (!("ggplot2" %in% names(installed.packages()[,"Package"]))) {
  install.packages("ggplot2")
  }
```

Then, we load the package:

```{r}
library(ggplot2)
```

## Task 1:

Visualize the relationship between life expectancy over time for Canada in the 1950s and 1960s using a line plot.

* Create a subset of `gapminder_5060` for Canada:
```{r}
ca_5060 = subset(gapminder_5060, country == "Canada")
```

* Create the plot:
```{r}
ggplot(data = ca_5060, aes(x = year, y = lifeExp)) + # creates the first layer
  geom_line()                                        # add line
```

### Task 1 - Stretch goal:

Add lines for Mexico and US in the 1950s and 1960s.

* Create a subset of `gapminder_5060` for Canada, Mexico, and US:
```{r}
countries = c("Canada", "Mexico", "United States")
northamer_5060 = subset(gapminder_5060, country %in% countries)
```

* Create the plot:
```{r}
ggplot(data = northamer_5060, 
       aes(x = year, y = lifeExp, colour = country)) + # add colour by country
  geom_line()                                          # add line 
```

## Task 2:

Visualize the relationship between GDP and life expectancy for countries in Europe in 1952 using a scatter plot.

* Create a subset of `gapminder_5060` for Europe in 1952:
```{r}
eu_52 = subset(gapminder_5060, continent == "Europe" & gapminder_5060$year == 1952)
```

* Create the plot:
```{r}
ggplot(data = eu_52, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()                                          # add points
```

### Task 2 - Stretch goal:

Add year 1967 in another color.

* Create a subset of `gapminder_5060` for Europe in 1952 and 1967:
```{r}
years = c(1952, 1967)
eu_5267 = subset(gapminder_5060, continent == "Europe" & (year %in% years))
```

* Create the plot:
```{r}
ggplot(data = eu_5267, aes(x = lifeExp, y = gdpPercap, colour = as.factor(year))) +
  geom_point()
```