---
title: "session2-template"
output: html_document
---


# Task 01

Gaol: Append the new data `gapminder-seventies-and-eighties.csv` and and `gapminder-nineties-and-later.csv` to your existing data set.

## Reading files

Reading in the data is quick and easy with the `read.csv` function.  
```{r}
gap_5060 <- read.csv('data/gapminder-5060.csv')
gap_7080 <- read.csv('data/gapminder-7080.csv')
gap_90plus <- read.csv('data/gapminder-90plus.csv')
```

There are a number of similarly named functions for reading files, such as  `read.delim` (tab-delimited files) and `read.table` (allows you to specify the separator), and there are similar third-party solutions for Excel spreadsheets and various other formats.

Once read in, the data are stored in a data structured called a "data frame". It's convenient to take a quick look at the data to make sure we know what it looks like. To do so, we'll use the `head` and `tail` functions which show the first and last lines of a data frame.

```{r}
head(gap_5060)
```


```{r}
tail(gap_5060)
```

When working in R markdown (via knitr) we can use the `kable` function to make a prettier version of the output from `head`.

```{r}
library(knitr)  # loads the knitr package, so we can use kable
kable(head(gap_90plus))
```


## Combining data frames

The functions `rbind` and `cbind` can be used to combine data frames. `rbind` combineds rows (such as when adding now observations to a data set), `cbind` combines by columns (useful when adding new variables to a data set).  

In the current context, we want to add new rows to our data set so we'll use `rbind`. One of the really convenient things about using `rbind` rather than trying to combine the spreadsheets in a program like Excel is that `rbind` matches columns by their names rather than locations, so a simple reordering of columns in one data frame versus another is very simply dealt with.


```{r}
gap_combined <- rbind(gap_5060, gap_7080, gap_90plus)
```


```{r}
kable(head(gap_combined))
```


```{r}
kable(tail(gap_combined))
```


# Task 02

Goal: Create line plots of life expectancy over time for Canada, Mexico, and the United States that run from 1952 to 2007.
- Stretch goal:  In the same plot, add similar line plots for Cambodia, China, and Japan and Egypt, South Africa, and Uganda.  Color countries with the same continent the same color.


```{r}
library(ggplot2)
```

Initial Goal
```{r}
north_america <- subset(gap_combined, country %in% c('Canada', 'Mexico', 'United States'))

ggplot(north_america, aes(x = year, y = lifeExp, color=country)) + geom_line() + xlab("Year") + ylab("Life Expectancy")
```

Stretch Goal:
```{r}

country_list <- c('Canada', 'Mexico', 'United States', 'Cambodia', 'China', 'Japan', 'Egypt', 'South Africa', 'Uganda')
gap_coi <- subset(gap_combined, country %in% country_list)

ggplot(gap_coi, aes(x = year, y = lifeExp, group=country, color=continent)) + geom_line() + xlab("Year") + ylab("Life Expectancy")

```

## Task 03



