---
title: "Code for exercises in Intro Session 1"
author: ''
date: ''
output:
  html_document:
    fig_height: 3
    fig_width: 6
---

This document contains the code for the exercises for this session. The code for the exercises  is also in `intro-01-template.Rmd`, which is the template RMarkdown document that will be shared with the participants at the beginning of Intro Session 2.

## Data:

Load the data:

```{r}
gapminder_5060 = read.csv("data/gapminder-5060.csv")
```

## Task 1:

Visualize the relationship between life expectancy over time for Canada in the 1950s and 1960s using a line plot.

* Create a subset of `gapminder_5060` for Canada:
```{r}
ca_5060 = gapminder_5060[gapminder_5060$country == "Canada", ]
```

* Create the scatterplot using `ggplot2`:
```{r}
library(ggplot2)
ggplot(data = ca_5060, aes(x = year, y = lifeExp)) +
  geom_point() +
  geom_line() 
```

### Task 1 - Stretch goal:

Add lines for Mexico and US in the 1950s and 1960s.

* Create a subset of `gapminder_5060` for Canada, Mexico, and US:
```{r}
countries = c("Canada", "Mexico", "United States")
northamer_5060 = gapminder_5060[gapminder_5060$country %in% countries, ]
```

* Create the scatterplot using `ggplot2`:
```{r}
library(ggplot2)
ggplot(data = northamer_5060, aes(x = year, y = lifeExp, colour = country)) +
  geom_line() +
  geom_point()
```

## Task 2:

Visualize the relationship between GDP and life expectancy for countries in Europe in 1952.

* Create a subset of `gapminder_5060` for Europe in 1952:
```{r}
eu_52 = gapminder_5060[gapminder_5060$continent == "Europe" & gapminder_5060$year == 1952, ]
```

* Create the plot:
```{r}
ggplot(data = eu_52, aes(x = lifeExp, y = gdpPercap)) +
  geom_line() +
  geom_point()
```

### Task 2 - Stretch goal:

Add year 1967 in another color.

* Create a subset of `gapminder_5060` for Europe in 1952 and 1967:
```{r}
eu_5267 = gapminder_5060[gapminder_5060$continent == "Europe" & (gapminder_5060$year == 1952 | gapminder_5060$year == 1967), ]
```

* Create the plot:
```{r}
ggplot(data = eu_5267, aes(x = lifeExp, y = gdpPercap, colour = as.factor(year))) +
  geom_line() +
  geom_point()
```